---
title: "ADA_Project"
author: "Caitlin O'Connell"
date: "9/25/2022"
output: html_document
---

## Load and Open packages
```{r}
pacman::p_load(openxlsx, ggplot2, tidyverse, table1, nhanesA, VIM, mice, lattice, odds.n.ends, lmtest, broom, car)
devtools::install_github("silentspringinstitute/RNHANES")
library(RNHANES)

# openxlsx used to write excel documents for export
# ggplot2 used to visually depict the association before running the analysis
# tidyverse used to do data management
# table1 used to generate descriptive statistics of cohort
# nhanesA and RHANES used to import the data from RNHANES
# VIM used for patterns within the missing data
# mice used for multiple imputation of missing data
# lattice used to look at imputed data
# odds.n.ends used to calculate odds ratios
# lmtest used to test for effect modification using the likelihood ratio test
# broom used to calculate cook's distance
# car used to test for multicolinearity
```

## Importing data sets from NHANES website (https://wwwn.cdc.gov/nchs/nhanes/continuousnhanes/default.aspx?BeginYear=2017)
```{r}
# physical activity questions with demographics
nhanes_pa <- nhanes_load_data("PAQ_J", "2017-2018", demographics = TRUE) 
# diabetes questionnaire
nhanes_diabetes <- nhanes_load_data("DIQ_J", "2017-2018") 
# weight questionnaire
nhanes_weight <- nhanes_load_data("WHQ_J", "2017-2018") 
# sleep questionnaire
nhanes_sleep <- nhanes_load_data("SLQ_J", "2017-2018") 
# health care utilization questionnaire
nhanes_health <- nhanes_load_data("HUQ_J", "2017-2018") 
# disability questionnaire
nhanes_disability <- nhanes_load_data("DLQ_J", "2017-2018") 
# occupation questionnaire
nhanes_occupation <- nhanes_load_data("OCQ_J", "2017-2018") 
# physical functioning questionnaire
nhanes_physfunc <- nhanes_load_data("PFQ_J", "2017-2018") 
# smoking questionnaire
nhanes_smoke <- nhanes_load_data("SMQ_J", "2017-2018") 

# A1C
nhanes_a1c <- nhanes_load_data("GHB_J", "2017-2018")
# fasting glucose
nhanes_glucose <- nhanes_load_data("GLU_J","2017-2018")
```

## Merging datasets together
```{r}
# merging diabetes with physical activity that includes demographics
merged_file <- merge(nhanes_diabetes,nhanes_pa, by = "SEQN")
# joining disability
merged_file <- left_join(merged_file, nhanes_disability, by="SEQN")
# adding health
merged_file <- left_join(merged_file, nhanes_health, by="SEQN")
# adding occupation
merged_file <- left_join(merged_file, nhanes_occupation, by="SEQN")
# adding physical functioning
merged_file <- left_join(merged_file, nhanes_physfunc, by="SEQN")
# adding sleep
merged_file <- left_join(merged_file, nhanes_sleep, by="SEQN")
# adding smoking
merged_file <- left_join(merged_file, nhanes_smoke, by="SEQN")
# adding weight
merged_file <- left_join(merged_file, nhanes_weight, by="SEQN")
```

## Selecting variables of interest
```{r}
# selecting variables needed for analysis: 
# participant id, SEQN
# gender (RIAGENDER), age (RIDAGEYR), race/ethnicity (RIDRETH3), education status (DMDEDUC2, DMDEDUC3), household income (INDHHIN2)
# diabetes status, DIQ010
# active transport variable, PAQ635
# physical activity variables, PAQ605, PAD645, PAQ650, PAQ655, PAD660, PAQ665,PAQ670, PAD675, 
# sedentary time, PAD680
# DLQ050 - difficulty walking 
# DLQ140 - How often do you feel depressed?
# HUQ010 - General Health condition
# OCQ670 - Overall work schedule past 3 months
# OCD390G - Kind of work you have done the longest
# SLD012 - Sleep hours - weekdays or workdays
# SMQ020 - Smoked at least 100 cigarettes in life
# SMQ040 - Do you now smoke cigarettes?
# WHD010 - Current self-reported height (inches)
# WHD020 - Current self-reported weight (pounds)

nhanes_small <- merged_file %>%
  select(SEQN, RIAGENDR, RIDAGEYR, RIDRETH3, DMDEDUC2, DMDEDUC3, INDHHIN2, DIQ010, PAQ605, PAQ635, PAD645, PAQ650, PAQ655, PAD660, PAQ665,PAQ670, PAD675, PAD680, DLQ050, DLQ140, OCQ670, OCD150, OCD390G, SLD012, SMQ020, SMQ040, WHD010, WHD020, HUQ010)
```

## Data Management
```{r}
nhanes_clean <- nhanes_small %>%
  mutate(diabetes = case_when(DIQ010 == 1 ~ 1,
                              DIQ010 == 2 ~ 0,
                              DIQ010 == 3 ~ 0,
                              DIQ010 == 7 ~ NA_real_,
                              DIQ010 == 9 ~ NA_real_)) %>%
  mutate(diabetes = factor(diabetes, c(0:1), c("No", "Yes"))) %>%
  mutate(act_transport = case_when(PAQ635 == 1 ~ 1,
                                   PAQ635 == 2 ~ 0)) %>%
  mutate(act_transport = factor(act_transport, c(0:1), c("No", "Yes"))) %>%
  mutate(sex = case_when(RIAGENDR == 1 ~ 0,
                         RIAGENDR == 2 ~ 1)) %>%
  mutate(sex = factor(sex, c(0:1), c("Male", "Female"))) %>%
  mutate(race = case_when(RIDRETH3 == 1 ~ 1,
                          RIDRETH3 == 2 ~ 2,
                          RIDRETH3 == 3 ~ 0,
                          RIDRETH3 == 4 ~ 3,
                          RIDRETH3 == 6 ~ 4,
                          RIDRETH3 == 7 ~ 5)) %>%
  mutate(race = factor(race, c(0:5), c("Non-Hispanic White","Mexican American","Other Hispanic","Non-Hispanic Black","Non-Hispanic Asian", "Other/Multi-Racial"))) %>%
  mutate(educat = case_when(DMDEDUC2 == 1 ~ 0,
                            DMDEDUC2 == 2 ~ 1,
                            DMDEDUC2 == 3 ~ 2,
                            DMDEDUC2 == 4 ~ 3,
                            DMDEDUC2 == 5 ~ 4,
                            DMDEDUC2 == 7 ~ NA_real_,
                            DMDEDUC2 == 9 ~ NA_real_,
                            DMDEDUC3 == 9 ~ 1,
                            DMDEDUC3 == 10 ~ 1,
                            DMDEDUC3 == 11 ~ 1,
                            DMDEDUC3 == 12 ~ 1,
                            DMDEDUC3 == 13 ~ 2,
                            DMDEDUC3 == 14 ~ 2,
                            DMDEDUC3 == 15 ~ 3,
                            DMDEDUC3 == 66 ~ 0,
                            DMDEDUC3 %in% c(0:8) ~ 0)) %>%
  mutate(educat = factor(educat, c(0:4), c("less than 9th grade","9th-12th grade (no diploma)","High school/GED","Some college","College or above"))) %>%
  mutate(income = case_when(INDHHIN2 == 15 ~ 0,
                            INDHHIN2 == 14 ~ 1,
                            INDHHIN2 == 13 ~ 8,
                            INDHHIN2 == 12 ~ NA_real_,
                            INDHHIN2 == 10 ~ 2,
                            INDHHIN2 == 9 ~ 3,
                            INDHHIN2 == 8 ~ 4,
                            INDHHIN2 == 7 ~ 5,
                            INDHHIN2 == 6 ~ 6,
                            INDHHIN2 == 5 ~ 7,
                            INDHHIN2 == 4 ~ 8,
                            INDHHIN2 == 3 ~ 8,
                            INDHHIN2 == 2 ~ 8,
                            INDHHIN2 == 1 ~ 8,
                            INDHHIN2 == 77 ~ NA_real_,
                            INDHHIN2 == 99 ~ NA_real_)) %>%
  mutate(income_underover = if_else(INDHHIN2 %in% c(1:4,13), 0, if_else(INDHHIN2 %in% c(5:12,14,15),1,NULL))) %>%
  mutate(income = factor(income, c(0:8),c("$100,000 and over",
                                          "$75,000 - $99,999",
                                          "$65,000 - $74,999",
                                          "$55,000 - $64,999",
                                          "$45,000 - $54,999",
                                          "$35,000 - $44,999",
                                          "$25,000 - $34,999",
                                          "$20,000 - $24,999",
                                          "Under $20,000"))) %>%
  mutate(health_status = case_when(HUQ010 == 1 ~ 0,
                                   HUQ010 == 2 ~ 1,
                                   HUQ010 == 3 ~ 2,
                                   HUQ010 == 4 ~ 3,
                                   HUQ010 == 5 ~ 4,
                                   HUQ010 == 7 ~ NA_real_,
                                   HUQ010 == 9 ~ NA_real_)) %>%
  mutate(health_status = factor(health_status, c(0:4), c("Excellent", "Very Good", "Good", "Fair", "Poor"))) %>%
  mutate(diff_walk = case_when(DLQ050 == 1 ~ 1,
                               DLQ050 == 2 ~ 0,
                               DLQ050 == 7 ~ NA_real_,
                               DLQ050 == 9 ~ NA_real_)) %>%
  mutate(diff_walk = factor(diff_walk, c(0:1), c("No", "Yes"))) %>%
  mutate(depress = case_when(DLQ140 == 1 ~ 0,
                             DLQ140 == 2 ~ 1,
                             DLQ140 == 3 ~ 2,
                             DLQ140 == 4 ~ 3,
                             DLQ140 == 5 ~ 4,
                             DLQ140 == 7 ~ NA_real_,
                             DLQ140 == 9 ~ NA_real_)) %>%
  mutate(depress = factor(depress, c(0:4), c("Daily","Weekly","Monthly", "A few times a year", "Never"))) %>%
  mutate(work_sch = case_when(OCQ670 == 1 ~ 0,
                              OCQ670 == 2 ~ 1,
                              OCQ670 == 3 ~ 2,
                              OCQ670 == 5 ~ 3,
                              OCD150 == 3 ~ 4,
                              OCD150 == 4 ~ 4,
                              OCQ670 == 7 ~ NA_real_,
                              OCQ670 == 9 ~ NA_real_)) %>%
  mutate(work_sch = factor(work_sch, c(0:4), c("Traditional (9to5)","Evenings or Nights","Early morning", "Variable", "Not working"))) %>%
  mutate(smoker = if_else(SMQ020 == 2, 0,
                           if_else(SMQ020 == 1 & SMQ040 == 1, 2,
                                   if_else(SMQ020 == 1 & SMQ040 == 2, 2,
                                           if_else(SMQ020 == 1 & SMQ040 == 3, 1, NULL))))) %>%
  mutate(smoker = factor(smoker, c(0:2), c("Nonsmoker", "Former", "Current"))) %>%
  mutate(mod_plus_vig = PAQ655 + PAQ670) %>%
  mutate(phys_act = if_else(PAQ650 == 2 & PAQ665 == 2, 0,
                              if_else(PAQ650 == 2 & PAQ665 == 1 & PAQ670 %in% c(1:2), 1,
                              if_else(PAQ650 == 2 & PAQ665 == 1 & PAQ670 %in% c(3:5), 2,
                              if_else(PAQ650 == 2 & PAQ665 == 1 & PAQ670 %in% c(6:7), 3,
                              if_else(PAQ650 == 1 & PAQ665 == 2 & PAQ655 %in% c(1:2), 1,
                              if_else(PAQ650 == 1 & PAQ665 == 2 & PAQ655 %in% c(3:5), 2,
                              if_else(PAQ650 == 1 & PAQ665 == 2 & PAQ655 %in% c(6:7), 3,
                                      if_else(PAQ650 == 1 & PAQ665 == 1 & mod_plus_vig %in% c(1:2), 1,
                                      if_else(PAQ650 == 1 & PAQ665 == 1 & mod_plus_vig %in% c(3:5), 2,
                                      if_else(PAQ650 == 1 & PAQ665 == 1 & mod_plus_vig %in% c(6:7), 3,
                                              if_else(PAQ650 == 1 & PAQ655 %in% c(1:2), 1,
                                              if_else(PAQ650 == 1 & PAQ655 %in% c(3:5), 2,
                                              if_else(PAQ650 == 1 & PAQ655 %in% c(6:7), 3, 
                                              if_else(PAQ665 == 1 & PAQ670 %in% c(1:2), 1,
                                              if_else(PAQ665 == 1 & PAQ670 %in% c(3:5), 2,
                                              if_else(PAQ665 == 1 & PAQ670 %in% c(6:7), 3,NULL))))))))))))))))) %>%
  mutate(phys_act = factor(phys_act, c(0:3), c("None","1 to 2 days", "3 to 5 days", "6 to 7 days"))) %>%
  mutate(PAD680 = na_if(PAD680, 9999)) %>%
  mutate(WHD010 = na_if(WHD010, 9999)) %>%
  mutate(WHD010 = na_if(WHD010, 7777)) %>%
  mutate(WHD020 = na_if(WHD020, 9999)) %>%
  mutate(WHD020 = na_if(WHD020, 7777)) %>%
  mutate(bmi = (WHD020/(WHD010*WHD010))*703) %>%
  mutate(sleep_time = case_when(SLD012 == 2 ~ 0,
                                SLD012 %in% c(3,3.5,4,4.5) ~ 0,
                                SLD012 %in% c(5,5.5,6,6.5) ~ 1,
                                SLD012 %in% c(7,7.5,8,8.5) ~ 2,
                                SLD012 %in% c(9,9.5,10,10.5) ~ 3,
                                SLD012 %in% c(11,11.5,12,12.5,13,13.5,14) ~ 4)) %>%
  mutate(sleep_time = factor(sleep_time, c(0:4), c("less than 4.5 hrs","5-6.5 hrs","7-8.5 hrs","9-10.5 hrs","11 hrs or more"))) %>%
  mutate(sed_time = PAD680) %>%
  rename(age = RIDAGEYR)

# checking the recoding
table(nhanes_clean$PAQ635, nhanes_clean$act_transport, useNA = "always")
table(nhanes_clean$DIQ010, nhanes_clean$diabetes, useNA = "always")
table(nhanes_clean$RIAGENDR, nhanes_clean$sex, useNA = "always")
table(nhanes_clean$RIDRETH3, nhanes_clean$race, useNA = "always")
table(nhanes_clean$DMDEDUC2, nhanes_clean$educat, useNA = "always")
table(nhanes_clean$DMDEDUC3, nhanes_clean$educat, useNA = "always")
table(nhanes_clean$INDHHIN2, nhanes_clean$income, useNA = "always")
table(nhanes_clean$INDHHIN2, nhanes_clean$income_underover, useNA = "always")
table(nhanes_clean$HUQ010, nhanes_clean$health_status, useNA = "always")
table(nhanes_clean$DLQ050, nhanes_clean$diff_walk, useNA = "always")
table(nhanes_clean$DLQ140, nhanes_clean$depress, useNA = "always")
table(nhanes_clean$OCQ670, nhanes_clean$work_sch, useNA = "always")
table(nhanes_clean$SMQ020, nhanes_clean$smoker, useNA = "always")
table(nhanes_clean$PAQ650, nhanes_clean$phys_act, useNA = "always")
table(nhanes_clean$PAQ655, nhanes_clean$phys_act, useNA = "always")
table(nhanes_clean$PAQ665, nhanes_clean$phys_act, useNA = "always")
table(nhanes_clean$PAQ670, nhanes_clean$phys_act, useNA = "always")
table(nhanes_clean$PAQ655, nhanes_clean$phys_act, useNA = "always")
table(nhanes_clean$SLD012, nhanes_clean$sleep_time, useNA = "always")

# data exploration
summary(nhanes_clean$WHD010)
summary(nhanes_clean$WHD020)
summary(nhanes_clean$bmi)
summary(nhanes_clean$sed_time)
summary(nhanes_clean$age)
```

## Defining Cohort
```{r}
# dropping those without diabetes status
# selecting clean variables for analysis
cohort <- nhanes_clean %>%
  drop_na(diabetes) %>%
  select(SEQN, age, sex, race, educat, income, health_status, diff_walk, work_sch, smoker, phys_act, bmi, sleep_time, sed_time, diabetes, act_transport)
  
```

## Table 1 of Descriptive Statistics
```{r}
label(cohort$age)<-"Age (years)"
label(cohort$sex)<-"Sex"
label(cohort$race)<-"Race"
label(cohort$educat)<-"Education Level"
label(cohort$income)<-"Income"
label(cohort$health_status)<-"Health Status"
label(cohort$diff_walk)<-"Has difficulty walking"
label(cohort$work_sch)<- "Work Schedule"
label(cohort$smoker)<-"Smoking Status"
label(cohort$phys_act)<- "Recreational Physical Activity Level"
label(cohort$bmi)<-"BMI (kg/m^2)"
label(cohort$sleep_time)<-"Weekday Sleep Amount"
label(cohort$sed_time)<-"Sedentary Time (minutes)"
label(cohort$act_transport)<-"Active Transportation"



table1(~ age + sex + race + educat + income + health_status + diff_walk + work_sch + smoker + phys_act + bmi + sleep_time + sed_time + act_transport | diabetes, data = cohort, render.continuous = c(. = 'Mean (SD)'))
```

## Dropping Missing Data
```{r}
# dropping NA's for all variables in dataset
cohort_complete <- cohort %>%
  drop_na(age, sex, race, income, health_status, diff_walk, work_sch, smoker, phys_act, bmi, sleep_time, sed_time)
```

## Barchart to visualize whether there is a difference in the active transportation by diabetes status.
```{r}
cohort_complete %>%
  group_by(act_transport, diabetes) %>%
  count() %>%
  group_by(act_transport) %>%
  mutate(perc.act.trans = 100*n/sum(n)) %>%
  ggplot(aes(x = act_transport,
             y = perc.act.trans,
             fill = diabetes))+
  geom_col(position = "dodge") + 
  theme_minimal() + 
  labs(y = "Percent in each group",
       x = "\nActive Transportation",
       fill = "Diabetes Status") + 
  ylim(0,100)
```


## Initial logistic regression model
```{r}
# non-imputed model
model1<- glm(diabetes ~ act_transport + age + sex + race + income + health_status + diff_walk + work_sch + smoker + phys_act + bmi + sleep_time + sed_time, data=cohort_complete, family="binomial")
summary(model1) # get log results

Df <- odds.n.ends(model1) # get OR results
r<-as.data.frame(Df$`Predictor odds ratios and 95% CI`)

sheets<-list("Results" = r)
write.xlsx(sheets, file="analysis_tables.xlsx", keepNA=TRUE, overwrite=TRUE, rowNames = TRUE)
```

### Testing Assumption of Linearity
#### Age
```{r}
cohort_complete <- cohort_complete %>%
  mutate(age.times.logage = age * log(age)) # create term to test linearity

model_age <- glm(diabetes ~ act_transport + age + sex + race + income + health_status + diff_walk + work_sch + smoker + phys_act + bmi + sleep_time + sed_time + age.times.logage, data=cohort_complete, family="binomial") 

summary(model_age)
```

The age.times.logage is significant (p<.05). Assumption not met for age. Age must be categorized.

#### BMI
```{r}
cohort_complete <- cohort_complete %>%
  mutate(bmi.times.logbmi = bmi * log(bmi)) # create term to test linearity

model_bmi <- glm(diabetes ~ act_transport + age + sex + race + income + health_status + diff_walk + work_sch + smoker + phys_act + bmi + sleep_time + sed_time + bmi.times.logbmi, data=cohort_complete, family="binomial") 

summary(model_bmi)
```

The bmi.times.logbmi is significant (p<.05). Assumption not met for bmi. BMI must be categorized.

#### Sedentary Time
```{r}
cohort_complete <- cohort_complete %>%
  mutate(sed_time.times.logsed =sed_time * log(sed_time)) # create term to test linearity

model_sed <- glm(diabetes ~ act_transport + age + sex + race + income + health_status + diff_walk + work_sch + smoker + phys_act + bmi + sleep_time + sed_time + sed_time.times.logsed, data=cohort_complete, family="binomial") 

summary(model_sed)
```

The sed_time.times.logsed is significant (p<.05). Assumption not met for sedentary time. Sedentary time must be categorized.

## Categorical Age, BMI, and Sedentary time
```{r}
# getting quantile cutoffs for sedentary time
quantile(cohort_complete$sed_time)

cohort_complete <- cohort_complete %>%
  mutate(bmi_cat = case_when(bmi > 0 & bmi < 18.5 ~ 0,
                             bmi >= 18.5 & bmi < 25 ~ 1,
                             bmi >= 25 & bmi < 30 ~ 2,
                             bmi >= 30 ~ 3),
         bmi_cat = factor(bmi_cat, 0:3, c("underweight", "normal", "overweight", "obese"))) %>%
  mutate(age_cat = case_when(age >= 18 & age < 30 ~ 0,
                             age >= 30 & age < 40 ~ 1,
                             age >= 40 & age < 50 ~ 2,
                             age >= 50 & age < 60 ~ 3,
                             age >= 60 & age < 70 ~ 4,
                             age >= 70 & age <= 80 ~ 5),
         age_cat = factor(age_cat, 0:5, c("18-29","30-39","40-49","50-59","60-69","70 and older"))) %>%
  mutate(sed_time_cat = case_when(sed_time > 0 & sed_time <= 180 ~ 0,
                                  sed_time > 180 & sed_time <= 300 ~ 1,
                                  sed_time > 300 & sed_time <= 480 ~ 2,
                                  sed_time > 480 & sed_time <= 1320 ~ 3),
         sed_time_cat = factor(sed_time_cat, 0:3, c("3 hours or less","3-5 hours","5-8 hours","greater than 8 hours")))

# checking to make sure recode worked
summary(cohort_complete$bmi_cat)
by(cohort_complete$bmi, cohort_complete$bmi_cat, summary)

summary(cohort_complete$age_cat)
by(cohort_complete$age, cohort_complete$age_cat, summary)

summary(cohort_complete$sed_time_cat)
by(cohort_complete$sed_time, cohort_complete$sed_time_cat, summary)
```

#### New Descriptives Table
```{r}
label(cohort_complete$age_cat)<-"Age category (years)"
label(cohort_complete$sex)<-"Sex"
label(cohort_complete$race)<-"Race"
label(cohort_complete$educat)<-"Education Level"
label(cohort_complete$income)<-"Income"
label(cohort_complete$health_status)<-"Health Status"
label(cohort_complete$diff_walk)<-"Has difficulty walking"
label(cohort_complete$work_sch)<- "Work Schedule"
label(cohort_complete$smoker)<-"Smoking Status"
label(cohort_complete$phys_act)<- "Recreational Physical Activity Level"
label(cohort_complete$bmi_cat)<-"BMI Category"
label(cohort_complete$sleep_time)<-"Weekday Sleep Amount"
label(cohort_complete$sed_time_cat)<-"Sedentary Time Category (hours)"
label(cohort_complete$act_transport)<-"Active Transportation"



table1(~ age_cat + sex + race + income + health_status + diff_walk + work_sch + smoker + phys_act + bmi_cat + sleep_time + sed_time_cat + act_transport | diabetes, data = cohort_complete, render.continuous = c(. = 'Mean (SD)'))
```

Those 18-29 that have diabetes only has 7 observations. Therefore, age categories need to collapse. Excellent health status in the diabetes "yes" column only has 6 observations. Therefore, health status needs collapsed. There are only 3 observations in the underweight category that have diabetes. Therefore, bmi categories should be collapsed. 

#### Collapsing categories for age_cat, health status, and bmi_cat due to few too observations
```{r}
cohort_complete <- cohort_complete %>%
  mutate(age_cat2 = case_when(age_cat %in% c("18-29","30-39") ~ 0,
                             age_cat == "40-49" ~ 1,
                             age_cat == "50-59" ~ 2,
                             age_cat == "60-69" ~ 3,
                             age_cat == "70 and older" ~ 4),
         age_cat2 = factor(age_cat2, c(0:4),c("18-39","40-49","50-59","60-69","70 and older"))) %>%
  mutate(bmi_cat2 = case_when(bmi_cat %in% c("underweight", "normal") ~ 0,
                             bmi_cat == "overweight" ~ 1,
                             bmi_cat == "obese" ~2),
         bmi_cat2 = factor(bmi_cat2, c(0:2), c("underweight/normal","overweight","obese"))) %>%
  mutate(health_status2 = case_when(health_status %in% c("Excellent","Very Good") ~ 0,
                                    health_status == "Good" ~ 1,
                                    health_status == "Fair" ~ 2,
                                    health_status == "Poor" ~ 3),
         health_status2 = factor(health_status2, c(0:3), c("Excellent/Very Good","Good","Fair","Poor")))

# checking recoding
table(cohort_complete $age_cat, cohort_complete $age_cat2, useNA = "always")
table(cohort_complete $bmi_cat, cohort_complete $bmi_cat2, useNA = "always")
table(cohort_complete $health_status, cohort_complete $health_status2, useNA = "always")
```

#### New Descriptives Table2
```{r}
label(cohort_complete$age_cat2)<-"Age category (years)"
label(cohort_complete$sex)<-"Sex"
label(cohort_complete$race)<-"Race"
label(cohort_complete$educat)<-"Education Level"
label(cohort_complete$income)<-"Income"
label(cohort_complete$health_status2)<-"Health Status"
label(cohort_complete$diff_walk)<-"Has difficulty walking"
label(cohort_complete$work_sch)<- "Work Schedule"
label(cohort_complete$smoker)<-"Smoking Status"
label(cohort_complete$phys_act)<- "Recreational Physical Activity Level"
label(cohort_complete$bmi_cat2)<-"BMI Category"
label(cohort_complete$sleep_time)<-"Weekday Sleep Amount"
label(cohort_complete$sed_time_cat)<-"Sedentary Time Category (hours)"
label(cohort_complete$act_transport)<-"Active Transportation"



table1(~ age_cat2 + sex + race + income + health_status2 + diff_walk + work_sch + smoker + phys_act + bmi_cat2 + sleep_time + sed_time_cat + act_transport | diabetes, data = cohort_complete, render.continuous = c(. = 'Mean (SD)'))
```

## New model with categorical age, bmi, and sedentary time
```{r}
# model with categories
model2<- glm(diabetes ~ act_transport + age_cat2 + sex + race + income + health_status2 + diff_walk + work_sch + smoker + phys_act + bmi_cat2 + sleep_time + sed_time_cat, data=cohort_complete, family="binomial")
summary(model2) # get log results

Df2 <- odds.n.ends(model2) # get OR results
r2<-as.data.frame(Df2$`Predictor odds ratios and 95% CI`)

sheets<-list("Inital" = r,
             "Categorical" = r2)
write.xlsx(sheets, file="analysis_tables.xlsx", keepNA=TRUE, overwrite=TRUE, rowNames = TRUE)

```

## Checking Multicolinearity Assumption
```{r}
vif(model2)
```

Meets multicolinearity assumption.

## Checking influential observations
```{r}
# Plot Cooks's Distance
plot(model2, which = 4, id.n = 3, col="red")

# calculating Cook's Distance
model2.data <- augment(model2) %>%  
  mutate(index = 1:n()) 

# looking at first couple of rows of data
head(model2.data)

# defines cutoff variable
cuttoff <- mean(model2.data$.cooksd)*3 
# looking at observations with cook's distance above cutoff
View(model2.data %>% filter(.cooksd>cuttoff))
```


## Effect Modification by Sex
```{r}
# model with interaction term between sex and active transport
model_em <- glm(diabetes ~ act_transport + age_cat2 + sex + race + income + health_status2 + diff_walk + work_sch + smoker + phys_act + bmi_cat2 + sleep_time + sed_time_cat + sex*act_transport, data=cohort_complete, family="binomial")

# comparing models
lrtest(model_em,model2)

```

The likelihood ratio test is not significant (p=0.2689). Therefore, there is no effect modification by sex. 

## Looking at fasting glucose 
```{r}
# adding glucose information
cohort_complete_glu <- left_join(cohort_complete,nhanes_glucose, by = "SEQN")

# filtering out those without diabetes and no fasting glucose value
diabetes_cohort <- cohort_complete_glu %>% filter(diabetes == "Yes") %>% drop_na(LBXGLU)

# calculating mean glucose by active transport status
by(diabetes_cohort$LBXGLU, diabetes_cohort$act_transport, mean)
by(diabetes_cohort$LBXGLU, diabetes_cohort$act_transport, median)

# filtering out those with diabetes and no fasting glucose
nondiabetes_cohort <- cohort_complete_glu %>% filter(diabetes == "No") %>% drop_na(LBXGLU)

# calculating mean glucose by active transport status
by(nondiabetes_cohort$LBXGLU, nondiabetes_cohort$act_transport, mean)
by(nondiabetes_cohort$LBXGLU, nondiabetes_cohort$act_transport, median)
```

## Looking at A1C 
```{r}
# adding a1c information
cohort_complete_a1c <- left_join(cohort_complete,nhanes_a1c, by = "SEQN")

# filtering out those without diabetes and no A1C information
diabetes_cohort <- cohort_complete_a1c %>% filter(diabetes == "Yes") %>% drop_na(LBXGH)

# calculating mean glucose by active transport status
by(diabetes_cohort$LBXGH, diabetes_cohort$act_transport, mean)
by(diabetes_cohort$LBXGH, diabetes_cohort$act_transport, median)

# filtering out those with diabetes and no A1C information
nondiabetes_cohort <- cohort_complete_a1c %>% filter(diabetes == "No") %>% drop_na(LBXGH)

# calculating mean glucose by active transport status
by(nondiabetes_cohort$LBXGH, nondiabetes_cohort$act_transport, mean)
by(nondiabetes_cohort$LBXGH, nondiabetes_cohort$act_transport, median)
```


## Multiple Imputation
```{r}
# calculating missing data percentage
(count(cohort) - count(cohort_complete))/count(cohort)

# creating dataframe for multiple imputation
# dropping those without diabetes status or active transport information
# selecting variables needed for multiple imputation
cohort_imp <- nhanes_clean %>%
  drop_na(diabetes, act_transport) %>%
  select(SEQN, age, sex, race, educat, income, health_status, diff_walk, work_sch, smoker, phys_act, WHD010, WHD020, sleep_time, sed_time, diabetes, act_transport, income_underover)

# looking at missing data pattern
md.pattern(cohort_imp, rotate.names=TRUE)
```

#### Perfomring the Multiple Imputation
```{r}
# performing multiple imputation
imp<-mice(cohort_imp, m=18, maxit=5, seed=219) 

# printing imputation details
imp

```
#### Checking Imputation results
```{r}
# checking variables that were imputed
imp$imp$educat
imp$imp$WHD010
imp$imp$WHD020
imp$imp$income
imp$imp$health_status
imp$imp$diff_walk
imp$imp$work_sch
imp$imp$phys_act
imp$imp$sleep_time
imp$imp$sed_time

# Checking distribution between imputed data and non-imputed data
# education
stripplot(x=imp, data=educat~.imp, jit=TRUE,  pch=20,  xlab="1=Non-missing, Imputation number") # You should see that the imputed data is randomly distributed amongst the non-imputed data. If not, one reason may be an error in the model that was used to impute a specific variable

# bmi
stripplot(x=imp, data=WHD010~.imp, jit=TRUE,  pch=20,  xlab="1=Non-missing, Imputation number")

#income
stripplot(x=imp, data=income~.imp, jit=TRUE,  pch=20,  xlab="1=Non-missing, Imputation number")

# health status
stripplot(x=imp, data=health_status~.imp, jit=TRUE,  pch=20,  xlab="1=Non-missing, Imputation number")

# difficulty walking
stripplot(x=imp, data=diff_walk~.imp, jit=TRUE,  pch=20,  xlab="1=Non-missing, Imputation number")

# work schedule
stripplot(x=imp, data=work_sch~.imp, jit=TRUE,  pch=20,  xlab="1=Non-missing, Imputation number")

# physical activity
stripplot(x=imp, data=phys_act~.imp, jit=TRUE,  pch=20,  xlab="1=Non-missing, Imputation number")

# sleep time
stripplot(x=imp, data=sleep_time~.imp, jit=TRUE,  pch=20,  xlab="1=Non-missing, Imputation number")

# sedentary time
stripplot(x=imp, data=sed_time~.imp, jit=TRUE,  pch=20,  xlab="1=Non-missing, Imputation number")
```


#### Creating Complete dataset
```{r}
longimp <- mice::complete(data=imp, action="long", include = TRUE)
```

#### Data Management
```{r}
longimp <- longimp %>%
  mutate(bmi = (WHD020/(WHD010*WHD010))*703) %>%
  mutate(bmi_cat2 = case_when(bmi > 0 & bmi < 25 ~ 0,
                             bmi >= 25 & bmi < 30 ~ 1,
                             bmi >= 30 ~ 2),
         bmi_cat2 = factor(bmi_cat2, 0:2, c("underweight/normal", "overweight", "obese"))) %>%
  mutate(age_cat2 = case_when(age >= 18 & age < 40 ~ 0,
                             age >= 40 & age < 50 ~ 1,
                             age >= 50 & age < 60 ~ 2,
                             age >= 60 & age < 70 ~ 3,
                             age >= 70 & age <= 80 ~ 4),
         age_cat2 = factor(age_cat2, 0:4, c("18-39","40-49","50-59","60-69","70 and older"))) %>%
  mutate(sed_time_cat = case_when(sed_time >= 0 & sed_time <= 180 ~ 0,
                                  sed_time > 180 & sed_time <= 300 ~ 1,
                                  sed_time > 300 & sed_time <= 480 ~ 2,
                                  sed_time > 480 & sed_time <= 1320 ~ 3),
         sed_time_cat = factor(sed_time_cat, 0:3, c("3 hours or less","3-5 hours","5-8 hours","greater than 8 hours"))) %>%
  mutate(health_status2 = case_when(health_status %in% c("Excellent","Very Good") ~ 0,
                                    health_status == "Good" ~ 1,
                                    health_status == "Fair" ~ 2,
                                    health_status == "Poor" ~ 3),
         health_status2 = factor(health_status2, c(0:3), c("Excellent/Very Good","Good","Fair","Poor")))

# checking to make sure recode worked
summary(longimp$bmi_cat2)
by(longimp$bmi, longimp$bmi_cat2, summary)

summary(longimp$age_cat2)
by(longimp$age, longimp$age_cat2, summary)

summary(longimp$sed_time_cat)
summary(longimp$sed_time)
by(longimp$sed_time, longimp$sed_time_cat, summary)

table(longimp$health_status, longimp$health_status2, useNA = "always")
```

#### Converting back to mids object
```{r}
imp2 <- as.mids(longimp)
```

#### Logistic Regression on Imputed Data
```{r}
model_imp <- with(imp2, glm(diabetes ~ act_transport + age_cat2 + sex + race + income + health_status2 + diff_walk + work_sch + smoker + phys_act + bmi_cat2 + sleep_time + sed_time_cat, family="binomial"))

summary(model_imp)

pool(model_imp) # pool regression coefficients and standard errors using pool function

# creating data frame with odds ratios and confidence intervals
DF_imp <- summary(pool(model_imp))
DF_imp$OR <- exp(DF_imp$estimate)
DF_imp$lower95 <- exp(DF_imp$estimate - 1.96*DF_imp$std.error)
DF_imp$upper95 <- exp(DF_imp$estimate + 1.96*DF_imp$std.error)
```

#### Exporting to Excel
```{r}
write.xlsx(DF_imp, file="imputed_analysis_tables.xlsx", keepNA=TRUE, overwrite=TRUE, rowNames = TRUE)
```

#### Effect Modification by sex
```{r}
# model with interaction term between sex and active transport
model_imp_em <- with(imp2, glm(diabetes ~ act_transport + age_cat2 + sex + race + income + health_status2 + diff_walk + work_sch + smoker + phys_act + bmi_cat2 + sleep_time + sed_time_cat + sex*act_transport, family="binomial"))

# comparing models with likelihood-ratio test statistic
D3(model_imp_em,model_imp)
```

The p-value of likelihood-ratio test statistic is non-significant (p=0.083). Therefore, no significant effect modification by sex. 
